---
title: "Exploring Line Lengths in R Packages"
author: "Michael Ingrisch"
date: "November 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Are line lengths in R more lognormal than in python?

Jake van der Plas recently analyzed the line length used in Python packages. He was inspired by an analysis of tweet lenghts on Twitter, which observed that English tweets ...

I regularly use Python as a general purpose programming language, for purposes like image analysis, sorting of files and other tasks. In R, on the other hand, I am much more a user than a programmer, with a focus on data analysis, modeling and visualization. Jake vdPs analysis inspired me to reproduce this work in R, with the purpose to analyze line lengths in R code. From this task, I hope:

1. To become more familiar with R as a programming language 

2. To investigate whether the lack of a character limit leads to a line length distribution which follows a lognormal distribution.


3. To observe whether there are style differences w.r.t to line length between the tidyverse and base R.


# Counting lines in the dplyr package

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

First, we need a function that reports lines and their lengths in an arbitrary R package. In python, this would be straigthforward (for me), in R, I lack foundations. Since I work within the tidyverse, this function should return a dataframe with the columns Package, Line and Length. After some base R exercises, I produced the following function:

```{r library.location}

file_to_df <- function(filename){
  # read a file line-by-line into a dataframe
  df <- as_data_frame(readr::read_lines(filename))
  df
}

line_lengths <- function(package){
  # this function reads lines and counts their length in all R scripts in a given directory
  # For now, the source code of the package needs to be downloaded manually

  # list all R files in path
  pkg.path = file.path("src",package)
  filelist = list.files(path = pkg.path, 
                        pattern = '.*\\.R$', # all files that end with .R
                        full.names = T, # for convenience: full path
                        recursive = T)

    # for each R file: read lines and count length
  #for loops are evil, we use purrr::map
  dfs <- purrr::map(filelist, file_to_df)  
  
  df = bind_rows(dfs) %>% ## merge dataframes
    mutate(Package = package) %>% # insert the package name, for convenience
    mutate(line_length = nchar(value)) # determine line length
  return(df)
}
```



With the function ``` line_lengths```, we can now proceed with the fun part and investigate the line lengths in dplyr:
```{r visualize.dplyr}

dplyr.lines <- line_lengths("dplyr") 

dplyr.lines %>% 
  ggplot(aes(x=line_length)) + 
  geom_histogram(binwidth = 1, color = "dark gray", fill="light gray")
```



WE observe to spikes, one for empty lines and one for lines with a length of 75. We investigate the latter:
```{r}
dplyr.lines %>% 
  filter(line_length==75) %>% select(value)
```
Apparently, comment lines are often dash-filled to a line length of 75.

Following Jvdp, we proceed with unique lines only, in order to remove boring boiler plate code>

```{r}
dplyr.lines %>% 
  distinct() %>% 
  ggplot(aes(x=line_length)) + 
  geom_histogram(binwidth = 1, color = "dark gray", fill="light gray")
```

This looks much more like the kind of distribution which we expect after having read Jake vdP's article. There appears to be a bump in line lengths around 65-70 characters, which may well result form personal preferences and editor window widths. Otherwise, the distribution looks fairly smooth and approximately lognormal.


```{r}
dplyr.lines %>% 
  filter(line_length==2) %>% select(value)
```
Most of these lines appear to be closing parentheses - this is something which is rarely observed in the python world.


# More packages
```{r visualize.ggplot}

ggplot.lines <- line_lengths("ggplot2") 

ggplot.lines %>% 
  distinct() %>% 
  ggplot(aes(x=line_length)) + 
  geom_histogram(binwidth = 1, color = "dark gray", fill="light gray")
```

```{r}

mlr.lines <- line_lengths("mlr")
df.all <- bind_rows(ggplot.lines, dplyr.lines, mlr.lines)

df.all %>% 
  distinct() %>% 
  ggplot(aes(x=line_length)) +  
  geom_histogram(binwidth = 1, color = "dark gray", fill="light gray") +
  facet_wrap(~Package, scales = "free_y")
```

# Fitting a lognormal distribution
 
```{r lognormal}
lognorm_model <- function(x, amplitude, mu, sigma){
  amplitude *dlnorm(x,mu,sigma)
  }
```
 
We need to figure out how to fit this function to our histograms. nls may be an appropriate function. Also, we need to extract the histograms from the ggplot object. Oh dear.



 